---
title: "p8131_hw5_jsg2145"
author: "Jared Garfinkel"
date: "3/7/2020"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(pscl)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

## Problem 1

```{r}
crab_df = read_table2("./data/HW5-crab.txt") %>% 
  janitor::clean_names()

```

This dataset is `r nrow(crab_df)` by `r ncol(crab_df)`. It gives information about female crabs like the average weight is `r mean(pull(crab_df, wt)) %>% round(digits = 2)`. Other variables like carapace width, color, and the condition of the spine. The response variable is the number of satellite male crabs around one's nest.

```{r}
crab_df %>% 
  ggplot(aes(x = sa)) +
  geom_histogram(binwidth = 1)
```


```{r}
M1 <- glm(sa ~ w, family = poisson(link = log), 
                 data = crab_df)
summary(M1)
```


The rate ratio of the number of satellite males around one's nest is `r coef(M1)["w"] %>% exp() %>%  round(digits = 2)`. This indicates that the number of satellites increases as the width of the female increases. Next, we check for over-dispersion.

```{r}
res.p1 = residuals(M1, type = "pearson", data = crab_df)
G1 = sum(res.p1^2)
pval = 1 - pchisq(G1, df = 170) # low p-value indicates a lack of fit
phi1 = G1 / (170)
phi1
M1$deviance/M1$df.residual
```

```{r}
plot(qnorm((173+1:173+0.5)/(2*173+1.125)),
     sort(abs(res.p1)), 
     xlab = 'Expected Half-Normal Order Stats',
     ylab='Ordered Abs Pearson Residuals')
abline(a=0, b=1)
abline(a=0, b=sqrt(phi1), lty=2)
```

There appears to be overdispersion indicating the data may not follow the theoretical distribution.

```{r}
M2 <- glm(sa ~ w + wt, 
          family = poisson(link = log), 
          data = crab_df)
summary(M2)
```


A second model is proposed including the width of the carapace and the weight of the females. In this model, the rate ratio is `r coef(M2)["w"] %>% exp() %>% round(digits = 2)` with each unit increase in width holding the weight constant. The rate ratio for male satellites is `r coef(M2)["wt"] %>% exp() %>% round(digits = 2)` for each unit increase in weight holding width constant.

```{r}
res.p2 = residuals(M2, type = "pearson", data = crab_df)
G2 = sum(res.p2^2)
pval = 1 - pchisq(G2, df = 170)
phi2 = G2 / (170)
phi2
M2$deviance/M2$df.residual

summary(M2,
        dispersion = phi2)
```

```{r}
plot(qnorm((173+1:173+0.5)/(2*173+1.125)),
     sort(abs(res.p2)), 
     xlab = 'Expected Half-Normal Order Stats',
     ylab='Ordered Abs Pearson Residuals')
abline(a=0, b=1)
abline(a=0, b=sqrt(phi2), lty=2)
```

This model also appears to show overdispersion. After adjusting for overdispersion, it appears the predictors become insignificant in the model. This indicates that the variance of the predictors may be due to something else, not association with the response variable.

```{r, results = "hide"}
test.stat = M1$deviance - M2$deviance # deviance (from original model fitting)
df = 171 - 170
res.p = residuals(M2, type = "pearson")  
res.p 
G = sum(res.p^2) # calc dispersion param based on larger model
phi = G / 170
F.stat = test.stat / (df * phi)
pval.test1 = 1 - pf(F.stat, df, 170)
pval.test1 # .113
```

A p-value above 0.05 (p = `r pval %>% round(digits = 3)`) means we do not reject the null, and therefore we accept the smaller model.

## Problem 2

```{r}
para_df = read_table2("./data/HW5-parasite.txt") %>% 
  janitor::clean_names() %>% 
  drop_na(c(length, intensity)) %>%
  select(sample, intensity, year, length, area) %>% 
  mutate(year = factor(year),
         area = factor(area))
```

After removing missing values in the variables of interest, this dataframe is `r nrow(para_df)` by `r ncol(para_df)`. It includes information about where the fish were caught (area), the sizes of the fish (length), and the year in which the fish were caught. The response variable is the intensity of parasites in each fish, which occur as follows:

```{r}
para_df %>% 
  ggplot(aes(x = intensity)) +
  geom_histogram(binwidth = 20)
```


```{r, results = "hide"}
skimr::skim(para_df)

para_df %>% 
  distinct(year)



```

```{r}
para.pois = glm(intensity ~ area + length + year, family = poisson(link = log), data = para_df)

summary(para.pois)
```

This model shows that the rate ratio of the intensity of parasites is `r coef(para.pois)["area2"] %>% exp() %>% round(digits = 3)` in area 2 compared to area 1 holding length and year constant.

The rate ratio of the intensity of parasites is `r coef(para.pois)["area3"] %>% exp() %>% round(digits = 3)` in area 3 compared to area 1 holding length and year constant.

The rate ratio of the intensity of parasites is `r coef(para.pois)["area4"] %>% exp() %>% round(digits = 3)` in area 4 compared to area 1 holding length and year constant.

The rate ratio of the intensity of parasites is `r coef(para.pois)["length"] %>% exp() %>% round(digits = 3)` for each unit increase in length holding area and year constant.

The rate ratio of the intensity of parasites is `r coef(para.pois)["year2000"] %>% exp() %>% round(digits = 3)` in the year 2000 compared to 1999 holding area and length constant.

The rate ratio of the intensity of parasites is `r coef(para.pois)["year2001"] %>% exp() %>% round(digits = 3)` in the year 2001 compared to 1999 holding area and length constant.

So, area 4 is much more likely to have fish with higher intensity parasites, while larger fish are more likely to have lower intensity of parasites.

```{r}
res.para = residuals(para.pois, 
                     type = "pearson",
                     data = para_df)
G.para = sum(res.para^2)
pval = 1 - pchisq(G.para, df = 1184)
phi.para = G.para / (1184)
phi.para
para.pois$deviance / para.pois$df.residual

summary(para.pois, dispersion = phi.para)
```
```{r}
plot(qnorm((1191 + 1:1191 + 0.5)/(2*1191 + 1.125)),
     sort(abs(res.para)),
     xlab = 'Expected Half-Normal Order Stats',
     ylab = 'Ordered Abs Pearson Residuals')
abline(a=0, b=1)
abline(a=0, b = sqrt(phi.para), lty=2)
```

It appears that the residuals follow a parabolic curve indicating that there is a mismatch between the model and the observed values.

```{r}
para.0 <- zeroinfl(intensity ~ area + year + length, data = para_df)
summary(para.0)
```

This model shows that the rate ratio of the intensity of parasites is `r coef(para.0)["count_area2"] %>% exp() %>% round(digits = 3)` in area 2 compared to area 1 holding length and year constant.

The rate ratio of the intensity of parasites is `r coef(para.0)["count_area3"] %>% exp() %>% round(digits = 3)` in area 3 compared to area 1 holding length and year constant.

The rate ratio of the intensity of parasites is `r coef(para.0)["count_area4"] %>% exp() %>% round(digits = 3)` in area 4 compared to area 1 holding length and year constant.

The rate ratio of the intensity of parasites is `r coef(para.0)["count_length"] %>% exp() %>% round(digits = 3)` for each unit increase in length holding area and year constant.

The rate ratio of the intensity of parasites is `r coef(para.0)["count_year2000"] %>% exp() %>% round(digits = 3)` in the year 2000 compared to 1999 holding area and length constant.

The rate ratio of the intensity of parasites is `r coef(para.0)["count_year2001"] %>% exp() %>% round(digits = 3)` in the year 2001 compared to 1999 holding area and length constant.

Fitting a model for zero inflation our parasites model shows that areas 2 and 3 are associated with higher intensity of parasites than area 1. On the other hand, higher length remains associated with lower intensity of parasites. So, fitting a model for zero inflation changed the association between area and intensity.
